### 2019-06-01 reviewresponse.tex
first structure

\documentclass[12pt]{article}

\usepackage[margin=2cm]{geometry}

\title{Response to Reviewers}
\author{Agent Programming in the Cognitive Era\\by Bordini, El Fallah Seghrouchni, Hindriks, Logan, Ricci}
\date{Submitted to JAAMAS 20th Anniversary Special Issue}

\newenvironment{reviews}{\bigskip\itshape}{\upshape\bigskip}
\newenvironment{response}{\bigskip\normalfont}{\bigskip}
\newcommand{\tbd}[1]{\textsf{\textbf{To Do:} #1}}

\begin{document}
\maketitle

We would like thank the reviewers for the excellent suggestions they made. Before we start with the more detailed response to the excellent specific points made by the reviewers, we would like to emphasise first that this is a paper in the VIEWPOINT category of JAAMAS submissions. It is \emph{not} a technical paper \emph{nor} a survey paper. It is not the purpose here to write a survey on Agent Oriented Programming, and in fact we don't have the space for the kind of literature review the reviewers would like us to do (with excellent suggestions, but we were only able to address a limited number of those suggestions). That survey section on Programming Languages specifically is there only to give a brief view of the history of our research community, acknowledging some of the papers that had historical impact in the area. The most important related work, however, are the ones where specific AI techniques have been incorporated into Agent programming platforms. We aim to have a thorough account of that, so if we missed any work on that area, we will be happy to add further references.

Below, we reproduce the reviewer comments (in italics) and provide our responses (in regular font).

\begin{reviews}
Dear Authors,

Thank you for submitting your manuscript to JAAMAS. After careful review, the paper has been found to need major revisions before it could be reconsidered for publication in the journal. Below you will find reviewer comments, which we hope are helpful to you in preparing future versions of your article. As we have discussed separately, this will be a viewpoint and the reviews below generally approach it in that context. I look forward to seeing the revision.

\begin{response}
Any response???
\end{response}

With our very best wishes,

--The Editors in Chief
======================

Reviewer \#1: The paper presents a state of art of agent programming for the last 20
years, showing how it can contribute for the development of future
intelligent systems. In particular, it stresses the implementations
of the BDI model and indicates possible enhancements on its sense,
plan and act phases, particularly using other AI techniques.

In my opinion, the subject discussed in the paper is completely adherent
to this special issue.

The paper is well written and easy to read. One major revision that
should be made, in case of acceptance, concerns the presentation of
references in the text. There are 2 different styles: (XX and YY,
year) and XX and YY (year). The first should be used as \\cite while
the second should be used as \\citet, when the authors are subject of
the phrase. In most of the paper, occurs the opposite, and some
phrases are difficult to understand. Some examples:
page 5, lines 41 and 42, page 6 line 19, page 13 line 32.

Although the core of the paper discusses important questions linking
AOP and AI, in my opinion there are some issues that need to be
addressed in the final version of the paper, before acceptance:

1. better (more precise)  definition of some core concepts used in the text

2. relation between AOP and other important aspects of AI (that
perhaps are mentioned in the text and not particularly stressed as
they should have been)

Regarding the first remark, the paper uses quite indistinctively the
following expressions as synonyms:
- BDI model
- BDI programming language / agent programming language
- BDI programming language / BDI platform / BDI architecture
- BDI agent / agent
- agent programming / multi-agent programming

Two examples of such indistinct uses:

In page 15, the title of section 4.2 is AI embedded into agents, but
in the content you mention "... integrated into the BDI architecture
... ". What about other types of agents, not BDI? I understand the
focus of the paper is BDI agents; hence you must change the title of
the section.

\begin{response}
Thanks, the title of the section was changed as suggested.
\end{response}

In page 17, in Discussion section, you say "... how AI can be
integrated into the BDI model...". I do not think that you have
shown this in the paper, I have seen integration to BDI agent
programming languages, not to the BDI model.

\begin{response}
We have corrected this to refer to the BDI model as used in agent programming languages specifically.
\end{response}

Most of the paper deals with BDI agent programming, considering a
single agent. In this sense, lines 29-36 in page 6, are *not*
implementations of the BDI programming model, although presented as
such in section 2.3. I understand the reference to EMAS (and its
predecessors) workshop but quite all the questions regarding
organizations, norms, environments were addressed by other
communities, like COIN for instance. Important references to the topic
of the paragraph, as Opera (Dignum), MOISE (Hubner) and Teams (Tambe)
should be mentioned.  Regarding the environment programming, is it
multi-agent programming or as a AI service, like you mention Cartago
in page 15? In other words, programming environments is part of multi-agent
programming or not? It is very confusing!

Regarding the 2nd remark, you do not make any mention to the current
and important issues of ethical and explainable AI. I believe that the
BDI approach can be extended regarding ethical issues (there are
published results about this, for instance, Dignum's and Boissier's
work) that deserves to be mentioned in such a vision of the future 20
years of AOP.

Some points that need to be made clearer:

page 2 - Title of section 2 is misleading, since except for the lines
mentioned above, you do not present multi-agent programming in this
section.

\begin{response}
We have changed the title as suggested.
\end{response}

page 3 - I see problems in some arrows in  Figure 1. Percepts can
generate goals directly? Who generate the intentions (no arrow
pointing to it)?

\begin{response}
Please note that the arrow says percepts/messages (the latter can generate goals directly. The link between the "interpreter" box and "intentions" is meant to cover also the creation of intentions.
\end{response}

page 7, line 19 - What does it mean "to be more appropriate to the
current context"?

\begin{response}
We tried to improve the sentence.
\end{response}

page 8, line 20/21, another issue you could have mentioned is time bounded
responses. This is precisely why the robotic community abandoned 
pure deliberative models. A reflection about this issue wrt BDI agents
must be done.

\begin{response}
\tbd{Brian, perhaps you could cite your work on real-time for agent languages to address this?}
\end{response}

page 8 - I particularly consider ML a subfield of AI. In several parts
of your text, you mention it in a different context from AI (see page
14, line 18). Please make a clear statement in the beginning of the
paper with respect to this concern.

\begin{response}
\tbd{Have we not done this? I think it was at some point in the text. Alessandro, perhaps you could check this when you handle your parts?}
\end{response}

page 9 - it would be nice to explain how an API should be integrated, in
general terms, to a BDI language. You mention this in lines 22-25, but
you don't explain in detail.

page 10 - Pantoja (2016) uses a result of Stabile ( BRACIS 2015) about perceptual
filtering in Jason, the text does not acknowledge this fact properly

page 11, lines 1-6, I'd add efficiency / bounded time response as a
challenge too

page 11, line 40-41, what is an intention plan?

\begin{response}
\tbd{Amal, I think you added the description to this work, this "intention plan" expression needs to be fixed or explained, and also note that a reviewer thought there was too much detail on this particular work.}
\end{response}

page 13, lines 38-39, what do you mean by an extended BDI
architecture? Integrating APIs as mentioned in my remark in page 9?
You should precise it better.

page 16, line 9, what does it mean "However implemented"?

\begin{response}
Regardless of the way in which it's been implemented
\end{response}

Some few typos:

page 7, line 18, ... in behaviour that *is* both more appropriated ...

page 9, line 22, *Their* paper introduces ....

page 11, line 35, languages. (double dot)

page 14, line 41, case, in which (insert blank)

page 16, line 13, built-in (sounds better)

\begin{response}
All the typos above were fixed, thanks.
\end{response}

Reviewer \#2: SUMMARY

The paper briefly argues that, despite the recent focus on machine learning, there is, and will continue to be, a place for agent programming. It goes on to survey the history of the field, highlighting both the contributions of the BDI model, and its shortcomings, as well as a range of work that aims to eliminate these shortcomings. Two approaches for integrating AI into agent programming are outlined. A broad trend is the move to add more feature into the language, so the level of programming is higher.

I would suggest that the authors improve the paper by:

1. Being more careful about which papers they select, ensuring broader more systematic coverage, and ensuring that they also indicate the level of maturity and significance

2. Expand the paper to have greater depth and nuance, especially in the discussion of the way forward (section 4)

3. Ideally, provide a useful roadmap, or challenges, that add something new to the discussions that have already taken place at recent EMAS and Dagstuhl events.

4. Please fix the bibliography, and please provide evidence for a range of claims (see below)

5. Either weaken the claims (in the cover letter) regarding the discussion of machine learning as not being the way forward, or expand this in the paper. Currently the discussion of why machine learning will not replace agent programming is quite brief, and is not the focus of the paper.

MAJOR

I started reading this paper with great interest, but ended up disappointed. The paper suffers from a number of significant flaws.

Firstly, it is quite brief - this means that none of the (important) topics are covered in the depth that they deserve. The survey of the state-of-the-art and the literature is brief and "sketchy", the

\begin{response}
There is a (purposely) brief survey of the history of AOP, as this is not meant to be a survey paper. Besides that section, other parts of the paper give examples of work that has combined AI techniques into AOP. Still, as a Viewpoint type of paper, those references are mainly there to give an ideia of the range of work done on such combination, but the main purpose of the paper is to put forward a vision for the future of AOP as facilitating the practical use of all sorts of AI techniques.
\end{response}

discussion of work aiming to alleviate the limitations of existing BDI languages is sketched out, and necessarily involves selecting only some work (which is acknowledged by the authors). And the discussion of future directions lacks depth and nuance. For instance, it fails to acknowledge the possible consequences of a richer and more powerful language, namely that the execution model can become more complex, so the programmer may find it harder to understand why a program is behaving the way that it is.  (This relates to discussion on page 16)

The paper also has a number of claims that are asserted without adequate justification or evidence (see detailed comments).

The paper also misses out on important issues. For example, the need for agent development tools to integrate with mainstream tools and platforms has been widely discussed in the community, but is not mentioned in the paper.

The selection of work to cover is in places odd. Both in terms of which papers are cited, and which are not, and also in terms of the level of detail. There are some papers cited (see below) that do not appear to be significant and influential work (and indeed, some papers cited are only 2-3 page extended abstracts!), whereas there is work that has not been cited that really should be. The level of detail of discussion is also oddly inconsistent: an important piece of work might receive a few sentences, while a specific and minor piece of work might get a whole paragraph. This issue could be mitigated by providing information on the maturity and level of development of different pieces of work.

One example of work that is missing completely relates to the move from single to multi agent systems, and the need to support multi-agent interactions.

\begin{response}
\tbd{I think we have done this but needs double checking.}
\end{response}

There is a whole community that looks at norms, and this is barely mentioned. There is also highly influential work on social commitments and their role in engineering agent systems. There is also a whole community that considers goal reasoning (cf. panel at the most recent EMAS).  There is also highly influential work on teams of agents (e.g. Milind Tambe's STEAM paper, and subsequent work on Machinetta)

\begin{response}
These are excellent pointers, of which we are of course familiar, but note that it would be an impossible task to discuss all the other areas of AAMAS which relate to AOP in one way or another. Our aim is to contribute to the JAAMAS anniversary special issue focusing on the EMAS community only, because we assume all other communities (like normative systems, social commitments, teamwork, etc.) will have their own papers. Although it's tempting to extend our paper since Tambe, Singh, and many others related to the points above have heavily influenced the workshops that formed the EMAS community.
\end{response}

I also have to comment that the selection of work appears to be biased towards the authors' own work. Of the 98 or so citations, almost half feature at least one of the authors of this paper as a co-author. This is very natural, since the authors are of course most familiar with their own work. The authors of this paper *are* key figures in the community, and have produced much work that is influential. However, this proportion strikes me as nonetheless being rather high.

\begin{response}
We completely agree with the reviewer and have tried to lower the number of self-citations. \tbd{We need to reduce significantly the self citations or increase the others!!!}
\end{response}

The selection of work is also somewhat myopic, in the sense that it focuses heavily on work associated with the EMAS community. I do not see this as a big issue, since the paper is effectively a review of this community's work, but it ought to be acknowledged explicitly.

\begin{response}
Thanks, we have explicitly acknowledged that in the paper (new paragraph at the end of Introduction).
\end{response}

Finally, please please please go through and carefully fix the bibliography! There is widely varying levels of detail in citations, more than one lower case "bdi", and two cases of the same paper being listed twice with slightly different details.

\begin{response}
Apologies for this, the paper was put together in a hurry to meet the special issue deadline. \tbd{Still TBD!!!!!}
\end{response}

DETAILED COMMENTS

(Page numbers refer to the number printed on the top of the page, and line numbers to the closest number printed on the side of the page)

Citations are often done incorrectly: using (AUTHOR, YEAR) when the citation is part of the sentence (see second paragraph of introduction), and, in many later cases, using AUTHOR (YEAR) when the citation is just an annotation to the sentence.

The discussion of Corea (2018) is difficult to follow without having read the original paper.

P2, line 26: "arguably the main contribution" - this is somewhat too strong. Perhaps weaken to "... one of the main"

\begin{response}
Done as suggested.
\end{response}

P5, L36: you claim the two most prominent BDI-based languages of the period were AgentSpeak and 3APL. This is debatable - and I would suggest that PRS certainly deserves a mention here!

\begin{response}
The claim was weakened and PRS referred to.
\end{response}

P6: (history) - one thing that would help is some indication of the maturity and use of different languages. There is a big difference between a language that was proposed in a short paper, implemented as a rough prototype, and used only by its author for a few small problems, and a language that was developed by a group over time, and was used for many applications. The selection of languages in the history is obviously limited (can't cover everything) but sometimes obscure minor languages are selected, and much more significant and well-developed languages are left out. For example, JIAC is not mentioned. Nor is JAM.

\begin{response}
Referring specifically to the section on the history of AOP, we did not identify which languages the reviewer considers obscure and minor. It is true we purposely omitted languages that are well known such as JAM or have been used for developing many applications like JIAC, because we restricted ourselves to a few languages from each "stage" of the history, citing only those that are widely agreed as having impacted the history of AOP. We don't mind adding a few more specific languages if the reviewer still disagrees with our choice.
\end{response}

P7 (key contributions): this page contains a number of claims that raised my eyebrows, and that need to be either toned down, or provided with some evidence (e.g. citations). These include:

- BDI is the dominant paradigm in agent programming - what if we look more broadly at the AI community, not just the EMAS community?

\begin{response}
We made it more clear that BDI is dominant among Agent Oriented Programming Languages, not the development/programming of anything that might be called a multi-agent system including both academia and industry.
\end{response}

Is this still a true statement? What if we consider the range of agent applications developed? There have been surveys (e.g. the one by Dignum \& Dignum [not cited, and really should be], and the more recent survey by Muller and Fischer [ditto]). From memory they show that most agent systems are not developed using BDI languages.

- End of second paragraph you claim that BDI languages can improve programmer productivity. Again, this is a strong claim. You might want to cite the AAMAS 2006 paper by Steve Benfield et al, which makes these claims from an industry perspective (but sadly does not provide adequate data to substantiate their claims).

- Lines 18-19 (p7): BDI agent are more robust in the face of failure - more robust than what? Evidence for this claim?  Similarly, a little later you claim that such context specific plans are more cumbersome to code in non-BDI languages. This is not argued, and is not immediately obvious.

- Line 31: a BDI agent is often able to recover (when a plan fails) - this is not quite true: it of course depends on the programmer having provided sufficient alternative plans.

- Line 36: BDI programs with the deliberation cycle provides "lower programmer effort" (again, citation needed - see Benfield)

- Line 46-48: BDI approach is critical to intelligibility .. and in particular Intentions make it much easier for end-users to understand what an agent is doing. It is not at all clear why this should be the case - please expand and explain. (Also top of p8, where you say it is straightforward to provide explanation)

- P8, L12: citation would help to support the argument that lack of transparency and predictability is an issue - there are lots of papers that argue this.

Page 11: the discussion of AgLOTOS is oddly detailed (compared to other work in this section). I immediately wondered if the author of the work was also an author of this paper (spoiler: yes, they are).

Page 12, Lines 6-13: this paragraph belongs in the next bit (AI planning)

Page 15: there have been various papers proposing modular agent architectures - from memory Emma Norling had a paper in the AAMAS blue skies track recently, and I think Virginia DIgnum also had a recent paper.

Page 16, L5-7: how might existing BDI concepts form a foundation for integrating AI capabilities?

Bibliography:
- Cardoso 2016 and 2017 - one is a local conference, the other an extended abstract. Should this work be selected to be cited in this paper?

\begin{response}
The extended abstract was changed to a full paper to appear in AAMAS-2019. The local conference has proceedings available in IEEE's DL but we deleted this self-citation to help with the statistics.
\end{response}

- Dastani 2008a\&b are the same paper twice

\begin{response}
Fixed, thanks.
\end{response}

- Panison \& Bordini 2017b- another extended abstract

\begin{response}
This reference was deleted too.
\end{response}

- Winikoff 2005a\&b - duplicate

\begin{response}
Fixed, thanks.
\end{response}

- Yao 2014 and 2016a - extended abstracts

SPECIFIC CORRECTIONS

P2, line 19 into BDI agent -> into the/a BDI agent
P2, line 22: A.I. - for consistency leave out the dots
P2, L49: "intention" -> "intentions"

P4, footnote 1: expand (and define) acronym APL

P5, L30 "all different programming paradigms" - too strong - suggest "many different" (if you want to keep it, then you need to argue that every single programming paradigm has been considered)
P5, L32: expand acronym AOP
P5, L43: "There is a number of resources" -> "There are a number of resources"

\tbd{This part of the text disappeared, was the reference to existing surveys moved elsewhere or just deleted? It may have been better to comment it out rather than delete.}

P6, L4: IRMA and PRS - citation?
P6, L6: expand acronyms ProMAS and DALT
P6, L16-18: unclear whether only atomic execution was formalised in 2APL, or also other Jason features. Sentence is grammatically awkward.
P6, L31: "of course" appears twice 
P6, L38: could perhaps also mention that EMAS arose from a Dagstuhl workshop

\tbd{The one above about Dagstuhl is still TO BE DONE!}

P10, L8: what is an "enthymeme"?

P14, L41: missing space before "In which"

\begin{response}
All the Specific Corrections above have been carried out, thanks.
\end{response}

Reviewer \#3: This paper provides a view point on the past and future role of the agent-oriented programming area in AI-based systems. I thank the authors for reflecting on this. I share their claim that AOP covers some requirements of intelligent systems that are not that well addressed by other AI approaches, such as ML. As such, AOP may, and probably should, be a part of a larger AI solution. I concur with the hypothesis that research and development in AOP is important and a special issue like this one would be seriously incomplete if there is no article on the past, present and future of AOP.

The team of authors is a very good representation of the work carried out in AOP and areas around it. The paper has a mixture of survey and future analysis/perspectives. Several of the main AOP approaches and systems are discussed, within a historical perspective rather than a technical one. The article also discusses what role AOP platforms can or should play in a larger AI solution, and what their advantages are wrt other AI techniques, in particular ML.

There are however three general comments I should state, in the hope to improve the paper so it better achieve its objectives. These comments are in the context of the aim of this special issue:

"The aim is not simply to review what has happened, but to understand what has been learned and what might need to be
done differently."

(Below, I list detailed examples that align with these main concerns, as well as "editorial" comments.)

===================
1) AUDIENCE. It was not clear to me what the expected audience of this article is. For what I could guess, it appears to be framed more for AOP researchers. The paper assumes familiarity with AOP aspects and names that would be, in my view, very hard to grasp for someone outside AOP. So, if this paper aims to educate, say, a ML researcher what AOP can bring to the table that they may not have,  then I think the paper needs a bit of reshaping. Things like "failure recovery" or how GOAL changes the focus from procedural plans to simple reactive rules, would be non-intelligible for a non AOP expert and there are no example, case studies discussed: which applications AOP are suit for? Section 2 is kind of redundant for an AOP person and not clear enough for an outsider (who would benefit from more concrete explanations and examples).

2) SURVEY. The survey aspect of the paper reads, in my view, much focused on the authors' work and a somehow "disorganized":

a. In some parts, it seems to suggest an historical perspective will be provided, which would suit the aim of the special issue. But, it starts basically with agents in the 2000's from the authors. For example, there is almost nothing on PRS, but in the end almost all the BDI systems are versions of PRS, much more modern of course. Very little is discussed about programming systems at the intentional stance, which would be part of the historical perspective.

b. It was hard to understand if the article is about agent-oriented programming or the specific BDI programming paradigm of AOP. In many places, it reads as the latter, including subtitles of sections. But then non-pure BDI languages are reviewed, like GOAL or Teleo-Reactive which are not per se core BDI frameworks; unless we take a broader view of BDI languages which I am totally fine but then one cannot ignore the huge work on Golog-like languages. Note that, while less SE "mainstream", the work on Golog-like in AOP has arguably done more connections with other fields, like belief representation, integration with planning or even decision theory (DTGolog).

I am aware that it is very hard to organize a survey on this area. The point is that it reads a bit of listing some somehow arbitrary list of work without a consistent overarching umbrella. The text flow is also hard to follow, in my view; in many parts reads  as an enumeration without much "aggregation".

I wonder if it is better, for both AOP researchers and outsiders, to give a more complete list of systems (table?), starting all the way from PRS say to even Java-based SARL (check that system, to me is one of best done today for "mainstream" programming) and set-up some aggreagated dimensions to compare/review them: reactive behavior (almost all), planning, plan-library based for non-functional requirements, advanced KR for beliefs, distributed computing (JADEX, SARL are strong), meta-level reasoning (e.g., meta-plans in JACK and PRS; reflection rules in 3APL), verification (Golog is very solid here), etc. Rather than a stream of citations with short text, I would have liked a more "aggregation" type of survey, a landscape. Then a table, easier to read, can be provided giving "ticks" or level of achievements to each, etc.

c. As a survey, it is missing some seminal references to refer to key issues, like Kowalski's for the the observe-plan-act cycle from the 80's (references from 2000's are given for that) and Cohen and Levesque's work on Intentions for the foundations of BDI agents. Also Golog-like languages assuming the paper is about AOP and not BDI only (Golog, ConGolog, IndiGolog, DTGOlog, RTGolog)

\begin{response}
\tbd{Could someone please add Kowalski's and C\&L references or fix response below?}
Thanks, we added references to those indeed seminal works. We already had a reference to Golog in Section 2.3 (which has a brief historical survey) and we added references to two of the Golog variants also making a point that most languages have had many variants but it's not our purpose to survey those here. In fact, please note that this is not a survey paper, it's in the Viewpoint category of JAAMAS.
\end{response}

3) FUTURE/REFLECTIONS. Here, I think the paper says very interesting things, but kind of "on the side" and late. As a reader, as an AOP expert and even more as an outsider, I would like to know very soon,  clearly, and precisely what AOP can bring to the table that other approaches would have a hard time. This includes things discussed (but late and on the side, again) about explainable AI/behavior, verifiable behavior, specification of non-functional requirements, and others such as reactive execution, integration of various AI technologies into an execution framework (note for example how simple SPARK at SRI played the integration platform role in the CALO project). Examples would help (e.g, in many real world there are business processes to follow).

Then, as an AOP researcher, I would like to know what are the top, say 5, specific problems that could be the key to success for AOP to be integrated with other AI solutions like ML. The paper is a bit "timid" on that and I think it would be nice to hear from this authors what they think are the top promising future directions that are not "incremental" steps. For example, enhancing BDI/AOP with built-in infrastructure to do the hard stuff, rather than letting the user deal with that: which ones and how?
===================

Overall I think this is a very good effort in putting together historical perspective, current context within AI, and future direction, and the team is very experienced to provide that story. I suspect it can be tweaked better as per above to achieve the message to either AOP practitioners or outsiders. Overall, the AOP researcher will be interested in reflecting and *aggregating* the key work done so far (more than enumeration) and understand what should come next, in particular in the overall AI context. On the other hand, an outsider (say from ML), needs to understand after reading the paper what is that AOP can contribute that his/her techniques find it hard.

I provide below some detailed comments, some even at editorial level and some that are examples of the three issues above. I apologize if they are not so "clean".

Page 2, Highlight (Orange):
Content: "Several large-scale organisations have begun to"
Comment: Would be nice to have Examples supporting this claim.

Page 2, Highlight (Yellow):
Content: "machines that sense, learn, reason, and interact with people in new ways to provide insight and advice."

Page 3, Line Drawing (Red)
Comment: More references, including a Cohen and Levesque in particular.

Page 3, Highlight (Orange):
Content: "(Rao and Georgeff, 1991)"

There are many cases where the reference is part of the sentence but it does not fit well and should be within (), unless you are referring to the authors themselves (e.g., Rao and Georgeff (1991) showed that....).

Page 4, Highlight (Orange):
Content: "'sense-plan-act' cycle"
Comment: I believe Kowalski is one of the first to refer to this cycle, would cite that in a a per like this one.

Page 5, Highlight (Orange):
Content: "soning over ontologies."
Comment: Or constraints ?

Page 6, Highlight (Orange):
Content: "of Golog (Lesp´erance et al., 1995)."
Comment: Ok but Golog was more offline planning. Later Indigolog is more "related" to BDI I guess...

Page 6, Highlight (Custom Color: \#bf00bf):
Content: "are survey"
Comment: there are also

Page 7, Highlight (Custom Color: \#bf00bf):
Content: "Dagstuhl Seminar Dix and Fisher"
Comment: Reads strange the name of the organisers like that

Page 7, Underline (Red):
Content: " based"

Page 7, Underline (Red):
Content: "based"

Page 7, Highlight (Custom Color: \#bf00bf):
Content: "was later"
Comment: were?

Page 7, Highlight (Custom Color: \#bf00bf):
Content: " Themain idea behind GOAL was to make the language simpler than existing languages,in particular providing rules that lead to the execution of individual actions rather thancourses of actions as in most languages which were typically influenced by the ideason reactive planning systems Georgeff and Lansky (1987a) from the AI literature."
Comment: Hard to parse. Also sections aims at giving historical overview but it starts discussions systems from year 2000, with PRS mentioned on the side. This is strange.

Page 7, Note (Green):
Since there are many proper names, would be nice to use another font for system names.

Page 7, Highlight (Orange):
Content: "and JACK Winikoff (2005b)."
Comment: Jack cam many years before and there are better refs, such as

Paolo Busetta, Ralph Ronnquist, Andrew Hodgson, and Andrew Lucas (1999) JACK Intelligent Agents - Components for Intelligent Agents in Java, AgentLink News, Issue 2.

Page 7, Highlight (Orange):
Content: "Java rather than creating new languages as such."
Comment: Why is this stated as a "negative" aspect with a "but"?. Isn't this a feature, in particular in the context of this paper about AOP impact on mainstream and AI? maybe I am misreading this, but ...

Page 7, Highlight (Custom Color: \#bf00bf):
Content: "(although of course the research strand started earlier of course)"
Comment: Of course twice

Page 7, Highlight (Custom Color: \#bf00bf):
Content: "the growing interest on"
Comment: that of

Page 8, Highlight (Orange):
Content: " (Georgeff et al., 1999). "
Comment: Quite old reference for this claim

Page 8, Highlight (Orange):
Content: "AI planning,"
Comment: First time mentioned, ref?

Page 8, Highlight (Orange):
Content: "(re)planning"
Comment: What do we mean by planning here? It's different from the previous use of the term. The BDI systems explained above do not do planning per se.

Page 8, Highlight (Orange):
Content: "backtracking"
Comment: This could be very confusing for an outsider, as backtracking is associated with search and planning. The failure recovery process is backtracking at the meta goal execution level, right? I would clarify this, as the idea is for a non AOP expert to read this, I would imagine.

Page 8, Highlight (Yellow):
Content: "learned, is the intelligibility of the resulting agent be- haviour by end users and stakeholders."
Comment: Very good!

Page 9, Highlight (Yellow):
Content: "The very high level of ab- straction (compared to conventional programming) also makes it easier to formally verify complex decision making, allowing us to give guarantees that an agent will be- have correctly in all circumstances (Fisher et al., 2013)."
Comment: Good. I think for this paper all these have to come before and more in the foreground

Page 9, Highlight (Orange):
Content: " are essentially limited to: - selecting pre-defined plans at run time based on the triggering event and theagent's current beliefs; and - some support for handling plan failure, e.g., aborting the plan in which the failureoccurred and trying another applicable plan to achieve the current subgoal. While these features are useful,"
Comment: I would enumerate what the systems do NOT do, rather than what they do. The text even recognised that these two are useful, so they are not limitations.

For example, cannot build new plans not programmed already, and have little support for the key selection functions, plans and intentions selection.

Page 9, Highlight (Orange):
Content: "AI or machine learning techniques,"
Comment: ML is an AI technique, right?

Page 10, Highlight (Orange):
Content: "Knowledge representation and reasoning"
Comment: Perfect, but this does. It seem to match the above limitations given.

Page 10, Highlight (Orange):
Content: "more expressive queries to the belief base,"
Comment: This is strange, how could it be possible if DLs are special types of FOL? How can they achieve more expressively than FOL? DL is about achieving enough expressivity with good complexity

Page 12, Highlight (Orange):
Content: "(Airiau et al., 2008)."
Comment: This is very preliminary work on that, Dhirendra Singh has much more developed approach and articles, including a JAAMAS paper.

Page 13, Highlight (Orange):
Content: "Also falling within this general approach are techniques based on HTN planning and search. For example, Sardi˜na et al. (2006) show how an HTN planner can be in- tegrated into a BDI architecture to find hierarchical decompositions of (agent) plans that avoid incorrect decisions at choice points, and so are less likely to fail during exe- cution. Similarly, Yao et al. (2014); Yao and Logan (2016); Yao et al. (2016b,a) show how techniques based on Monte-Carlo Tree Search can be used to select plans that minimise conflicts between intentions, achieve goals by their deadlines and recover fromfailures."
Comment: It is very strange to have these works listed under machine learning, they do no learning but planning, which is the heading just after. Very confusing to me.

Instead of using the techniques as headings, I would use the problem in AOP that are trying to address. For example, "smarter plan selection" or "synthesis of new plans", etc.

Page 13, Highlight (Orange):
Content: "AI planning"
Comment: This is one of the areas that are important in plugging BDI with other core AI areas, like automated planning. To me the explanation given below is extremely short and does not provide much to the outsider of AOP. What planner and technique is used for jade or Jason for example. Also the work of de Silva and Rao K, is missing, using first principle planning as "macro" for user intent, an extremely important point for BDI paradigm. Also, I believe Felipe M. Has done work in FPP for AgentSpeak.

Page 14, Highlight (Orange):
Content: "to avoid conflicts."
Comment: In my view, the. On AOP expert will not understand why this is fundamental issue, and how is this addressed in existing systems (or how it is not!)

Page 14, Highlight (Orange):
Content: "in Nilsson's Teleo-Reactive rule based agent programming lan- guage Nilsson (2001)."
Comment: I think it's great that TR is mentioned, but this opens the question why there are other very nice AOP approaches not reviewed. In mainstream agent SE I can think of SARL, one of the best I have seen lately. True, to core BDI but since TR and even GOAL are listed, I am confused which was the "principle" to select some. And also of course all the Golog like languages. True, not core BDI but the again same argument...

Page 14, Highlight (Orange):
Content: "integrating AI and ML into BDI-based"
Comment: And planning and constraints and other ai fields? The works above are not just about ML integrations so this sentences reads not aligned.

Page 15, Highlight (Orange):
Content: "the BDI archi- tecture and cycle."
Comment: See, here it seems the paper is only conceNed about core BDI

Page 16, Highlight (Orange):
Content: "EIS"
Comment: Can we spell?

Page 16, Highlight (Orange):
Content: "In this case,"
Comment: Which case? Section just started.

Page 16, Highlight (Orange):
Content: "PRS"
Comment: Note neither PRS nor dMars have been discussed, even on the historical content.

Page 17, Highlight (Yellow):
Content: "However implemented, the aimof this approach is to raise the level of abstraction of agent programming, by increasing the basic 'competence' of the agent language or platform. The agent developer no longer has to encode reasoning such as plan failure recovery, plan selection, and so forth in the agent's plans and/or by configuring the the agent increases, the role of the agent developer will change fromprogramming ex- actly what the agent will do in all situations, to providing more strategic information, heuristics, advice, social knowledge (e.g., norms), etc. about what the agent should (or should not) do in a given situation, leaving the details of the implementation of the strategy to the agent."
Comment: Exactly! This is fundamental point I believe for this paper and is left too late and underdeveloped. For example, how can we use statistical learning for this? Can we?

Page 17, Highlight (Orange):
Content: "at least for the foreseeable future."
Comment: Reads vague and doubtful, as if you don't even believe it much...

Page 17, Highlight (Yellow):
Content: "the critical aspects of the agent's behaviour are required to fallwithin a particular envelope or follow a particular pattern. "
Comment: This I think is also important and must be stated much before, clear, and with examples (how a plane is flew or business processes)

Page 18, Highlight (Yellow):
Content: "In its current form, it offers significantadvantages of intelligibility and verifyability that are difficult to replicate using only"

Page 28, Highlight (Orange):
Content: "The claim that future intelligent systems will be trained using machine learn- ing techniques rather than developed by human programmers has implications for ologies and tools for programming agents."
Comment: I don't understand the claim here. Is it that ml will for sure be used as such or that doing so may have not so great advantages? In any case, the question is about why this viewpoint is topical now. It seems because success in ML makes many believe KR - programming approaches are not or will not be needed, and you want to argue against that proposition?

Page 28, Highlight (Orange):
Content: "We provide a brief survey of the last 20 years of agent oriented programming and agent development platforms."
Comment: This is my major doubt. Are we after a survey paper?

Page 29, Highlight (Orange):
Content: "The (Hindriks 2014) paper traces the development of the agent oriented ap- proach to programming from the 1990s to the present day."
Comment: Ok but this is a big overlap with this paper, which it's majority is a survey, as far as I can see

\end{reviews}
\end{document}

### 2019-06-01 conclusion.tex
first structure

%% RHB: OK to change this title as below?
%\section{Conclusion and Future Work}
\section{Conclusion}
\label{sec:conclusion}

\todo{OLD, TO BE REWORKED}

In this paper, we introduced \aser, a novel extension of the classical
{\asl} language. The language provides encapsulation for agent
goals, which clearly improves legibility and reusability of AgentSpeak
code. Furthermore, the new language improves some of the shortcomings
of AgentSpeak in regards to goal orientation and declarative goals by
ensuring that all reactive plans are also associated with general
goals, providing a ``goal condition'' which means goals can be still
active even though presently there is no action for the agent to take
towards that goal, and allowing external events (i.e., reactions to
changes in beliefs) to trigger various plans, for all the goals it
might be relevant.  We formalised the main changes required in the
existing formal semantics of AgentSpeak and experimentally evaluated
an interpreter for \aser\ implemented on top of the ASTRA platform.

%% RHB: OK to mention the multi-agent contest?

% \begin{itemize}
% \item refining current ASTRA implementation, that will allow to refine the evaluation
% \item Jason extension implementing {\aser}, and compare it with the ASTRA one
% \item stress {\aser} in practice 
% \item think about the tools: how existing IDE and tools can be extended to provide functionalities exploiting the new features
% \item investigate how the extension either improve or not the possibility to formally analyse the correctness properties
% \item ...
% \end{itemize}

### 2019-06-01 evaluation-old.tex
first structure

\section{Discussion with Practical Examples and Cases}
\label{sec:discussion}

\todo{OLD EVALUATION SECTION}

The new language has been evaluated through a prototype implementation that builds on 
the ASTRA~\cite{DBLP:conf/prima/CollierRL15} language. This prototype, which is compatible
with pure ASTRA code, is used to reimplement part of an existing solution for the Minority 
Game~\cite{moro2004minority}. Interpreter cycle timings are then captured to develop an initial
understanding of how the new language performs.

{\small
\begin{verbatim}
g-plan +!g() : c <: gc { 
  body {
    // main plan body goes here...
    a; b; !g1(); !!g2();
  }

  /* e-plans */
  rule +e1 : c1 { b1; }
  rule +e2 : c2 { b2; }
}
\end{verbatim}}

As can be seen in the  snippet of code above, the implementation of the {\aser} prototype is 
syntactically different to the examples provided in Section~\ref{sec:proposal}. This reflects 
the syntactic approach adopted in ASTRA. It does not impact the associated semantics.

\subsection{Minority Game}
\label{minority}
The Minority Game is a well-known model for evaluating collective behaviour of agents competing 
for finite resources~\cite{moro2004minority}. In a nutshell, the game involves an odd number of
agents competing for a resource over a series of rounds. For a round, each agent makes a binary
decision (yes/no). At the end of the round, the bids are counted, and the agents that are in the
minority win. The game has been applied mainly in the areas such as modelling financial markets 
\cite{challet2013minority} and traffic simulation \cite{chmura2004minority}.

To provide an initial evaluation of {\aser}, we adapt an existing ASTRA-based implementation of 
the Minority Game (MG). As is shown in Figure~\ref{fig:mgagents}, the implementation consists of
3 types of agent: the \emph{compere} agent, who is responsible for managing the game (starting 
rounds, evaluating bids, ...); the \emph{player} agent, who implements a set of MG strategies; 
and the \emph{main} agent, which is responsible for configuring the game (creating and configuring
the compere and player agents). Interaction between the Compere and the Players is through a shared 
game-board artifact.

\begin{figure}[!tbh]
\centering
\includegraphics[height=2in, width=2.5in]{figs/mg.png}
\caption{Minority Game Agent Architecture}
\label{fig:mgagents}
\end{figure}

The existing implementation currently consists of 10 plans. Two of the plans are generic and are
used for configuration. Another plan implements the bidding behaviour, delegating to a 
\verb|!select_bid(...)| sub-goal that can be contextually selected based on the current strategy.
Of the remaining plans: four relate to the best play strategy; one relates to a random strategy;
and two relate to a sticky strategy.

In the {\aser} implementation, we reduce this to 3 g-plans --- one for each of the strategies --- 
and one traditional plan to handle initialization. The code below consists of the initialisation 
plan and a single g-plan representating the revised implementation of the best play strategy.

In the code below, variable \verb|A| represents the game board, which
is implemented as an ASTRA module~\cite{DBLP:conf/prima/CollierRL15}. Modules allow the developer to
create custom actions, sensors, (logical) terms, (logical) formulae,
and events. For example, \verb|A.join()| is a custom action to allow
the agent to join the game board; \verb|A.match(history, hist)| is a
custom term that returns the length of the longest common suffix of the
\verb|history| and \verb|hist| lists; and
\verb|$A.event(...)| is a custom event that is generated when the
Compere agent starts a new round.

The initialisation plan selects the game strategy to be employed by the player. This information
 is passed via the \verb|main(...)| goal; the ASTRA equivalent of a
 main method. When the interpreter handles
the event corresponding to this goal, it spawns a new intention to achieve the \verb|win(...)| 
goal (spawning is like forking in multi-processing and is denoted the double exclamation mark).

When the chosen strategy is \emph{bestplay}, the agent achieves the goal by selecting 
the \verb|!win("bestplay", ...)| g-plan. Upon adoption, the body of this plan is executed, which 
in this case creates a random set of strategies that are to be used by the agent to select its 
bid. This involves adopting the \verb|!setup_tactic()| subgoal which is handled by the first two 
rules within the g-plan. The start of a round is modelled through the custom event 
\verb|$A.event("play", [])|. The rule that is selected to handle this event chooses the best 
strategy from the set of strategies available to the agent. Finally, the end of a round is modelled 
through the custom event \verb|$A.event("winner", [int bid])|. The rule that is selected to handle 
this event updates the score for all strategies that lead to \verb|bid| being selected.

{\small
\begin{verbatim}
rule +!main([string strategy, list config]) {
  !!win(strategy, config); A.join();
}

g-plan +!win("bestplay", [int t, int h]) <: false {
  body { !setup_tactic(t); }

  rule +!setup_tactic(0) {}
  rule +!setup_tactic(int t2) {
    list hist = [];

    // construct a random history
    int i=0;
    while (i<h) {
      hist = hist + [M.randomInt() % 2];
      i++;
    }
    +strategy(t2, hist, M.randomInt() % 2);
    +score(t2, 0);

    !setup_tactic(t2-1);
  }

  rule $A.event("play", []) { 
    list history = A.results();

    int max_len = -1; int max_choice = -1;
    int max_score = -1;
    foreach (strategy(int s, list hist, int c) 
              & score(s, int sc)) {
      int len = A.match(history, hist);
      if ((len > max_len) | 
          ((len == max_len) & (max_score < sc))) {
        max_choice = c; max_score = sc;
        max_len = len;
      }
    }
    A.bid(max_choice);
  }

  rule $A.event("winner", [int bid]) {
    foreach (strategy(int s, list hist, bid) 
           & score(s, int sc)) {
      -+score(s, sc+1);
    }
  }
}
\end{verbatim}}

Some interesting observations from adapting the code were: (i) the compexity of the plan contexts 
were simplified when using g-plans because the g-plan itself provided some of the context; (ii) 
the number of arguments passed as parameters was reduced, again because the scope of the parameters 
of the g-plan was the plan body \emph{and} all of its sub-plans; (iii) the total number of rules 
under consideration on each iteration was significantly less because only the rules within an 
active g-plan were considered by the agent --- for example, in the original implementation, there 
was always at least one applicable \verb|+!select_bid()| plan per strategy, whilst in the new 
implementation only the plans related to the current strategy were applicable.

\subsection{Preliminary Analysis}
\label{performance}

We evaluate the prototype by comparing the {\aser} and ASTRA implementations. Initially,
we compared interpreter cycle execution time and the number of iterations based on a
single configuration of the MG with 29 players and 1000 rounds. For our results, we averaged
the values across all 29 players and repeated the experiment 5 times.

\begin{table}[]
\centering
\caption{Comparing ASTRA and {\aser}}
\label{comparison}
\begin{tabular}{lll}
                            & ASER    & ASTRA  \\
Cycle Time (ms)             & 0.0017  & 0.0036 \\
Cycles                      & 293,772 & 62,880 \\
Elapsed Execution Time (ms) & 495.22  & 229.29 \\
Unix (timed)                & 16s     & 15s   
\end{tabular}
\end{table}

Results of our initial comparison can be found in Table~\ref{comparison}. The difference
in the number of cycles is due primarily to the scheduling algorithm used by ASTRA, which
suspends agents that have no sensors (perceptors) and whose event and intention queues are empty.
The impact of this is that ASTRA is generally more efficient than {\aser}. Due to the small but 
consistent difference in performance between the unix timing of the two experiments, we then 
explored how increasing the number of rounds affected performance. Results for this are shown 
in Table~\ref{rounds}.

\begin{table}[]
\centering
\caption{ASTRA vs {\aser} Performance}
\label{rounds}
\begin{tabular}{llllll}
           & 1000   & 2000   & 3000   & 4000   & 10000   \\
ASER (s)   & 15.514 & 30.251 & 44.202 & 57.736 & 140.059 \\
ASTRA (s)  & 14.893 & 28.232 & 42.168 & 56.453 & 137.379 \\
Diff. (\%) & 4.2\%  & 6.8\%  & 4.8\%  & 2.3\%  & 2.0\%  
\end{tabular}
\end{table}

This second table shows that {\aser} has a small impact on performance, but it is 
almost linear in this case (ASTRA shows a performance improvement of between 2-6\%).

While this is not a thorough evaluation of {\aser}, it does hint that the use of
goal conditions does not significantly impact the performance of the language. It 
must also be noted that the prototype implementation is not as mature as the ASTRA
implementation --- interpreter optimisations could further reduce the difference 
in performance.

### 2019-06-01 formalisation.tex
first structure

\section{Formal Semantics}
\label{sec:formalisation}

%% First update reasoning cycle states...

Briefly, these are the changes in the semantics of AgentSpeak(L)
that are needed to accomodate the new features of AgentSpeaker(ER):

\begin{itemize}
\item in the beginning of the reasoning cycle, check all
  goal-conditions starting from the bottom of each intention stack and
  remove all finished intentions;

\item if an event is \emph{external}, try to trigger one
  applicable plan for each intention, starting from the
  top of the intention stack (that is, the most specific plan in the
  respective g-plan tree);

\item if an event is \emph{internal}, search for an applicable plan on
  the path to the root of the g-plan tree, that is, starting from the
  most specific plans in the g-plan tree (as in the item above but now
  there is only one intention to refer to); this does not change the
  semantics per se, only the way in which we look for relevant
  applicable plans.
\end{itemize}

We start by updating the well-known reasoning cycle~\cite{bordini:07}
to adapt it for the initial stage of checking all goal-conditions to
remove the intentions that should no longer be active. Furthermore,
note that even though this is computationally costly, it is necessary
because not doing it at every reasoning cycle may imply the agent
missing the moment where the goal-condition became true (hence an
intention needing to be deactivated); it is also worth the
computational burden in as much as it has the various practical
programming advantages we pointed out earlier in this paper.

%% RHB: I removed this text from above, but I guess it should be in
%% the initial sections
% Recall that this can be because the goal has been achieved, has become
% impossible to achieve, or the motivation why it was adopted no longer
% applies.

The required stage for checking goal-conditions is included in the
existing $\ClrInt$ stage (which previously only removed empty
intentions) except it is now moved to the beginning of the reasoning
cycle (to ensure nothing in the reasoning cycle is done under the
assumption a deactivated intention is still active), just after the
$\ProcMsg$ stage (as the information just received from other agents
might be useful in checking for goals to be deactivated). The clearing of
intentions used to be the last part of the reasoning cycle, but
because there are no other dependencies between the first and last
stages, $\ClrInt$ might as well be done at the beginning rather than
the end. The slightly changed reasoning cycle is shown in
Figure~\ref{fig:rcaser}.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width=\linewidth]{figs/ASERrc.pdf}
    \caption{The AgentSpeak(ER) Reasoning Cycle}
    \label{fig:rcaser}
  \end{center}
\end{figure}

The $\ClrInt$ stage also needs new semantic rules. In fact, the
previous 3 \rn{ClrInt} rules (see~\cite[p~212]{bordini:07}) are no
longer needed, as empty intentions are not to be removed from the set
of intentions, unless their goal condition becomes true. To facilitate
the presentation of the new rules, we define a new auxiliary function
$\GCOND$, which given a g-plan simply returns the goal-condition
component of that plan, as follows. Let $p=$``\texttt{+!g : c <: gc
  \{<- a.\}}'', then $\GCOND(p)=\texttt{gc}$, more specifically the
logical formula coded by \texttt{gc}. Also, we denote by
$[p_1,...,p_n]$ an intention stack with plan $p_n$ at its top.

The new rule \rn{ClrInt$_1$} is as shown below, and rule
\rn{ClrInt$_1$} is not shown because it simply causes the transition
$\CFG{\ClrInt} \trans \CFG{\SelEv}$ in case the negation of the
precondition of \rn{ClrInt$_1$} holds. The rule below essentially
removes from intentions the bottom-most g-plan for which its
goal-condition now follows from the state of the belief base. The
intuition is that if goal $\texttt{!g1}$ required a subgoal
$\mathtt{!g2}$ (a plan for which was pushed on top of the plan for the
former one in the intention stack) and $\texttt{!g1}$ is no longer
active, that whole part of the intention stack above it needs to be
removed together with it.

\infrule[ClrInt$_1$] {
        i \in \CI \qquad i=[p_1,\dots,p_n] \\
        \AGBELS \models \GCOND(p_j) \mbox{ for some } j, 1\leq j\leq n
}
%------------------------------------------------
{   \CFG{\ClrInt}  \trans  \CFGcp{\ClrInt} \\[1.2mm]  
\begin{array}{llcl}
  \mbox{\emph{where:}\quad}
     & \multicolumn{3}{l}{\mbox{$j$ is the least number in [1..n]
       s.t.}}\\
     & & & \AGBELS \models \GCOND(p_j)\\
     & \CIli & = & \CI \setminus \{i\} \cup \{[p_1,\ldots,p_{j-1}]\} \\
\end{array}
}

The auxiliary functions $\RELPLANS(\PLANS,\TE)$
$\APPLPLANS(\PLANS,\TE)$ (see Definitions~10.2 and~10.3
in~\cite{bordini:07}) can be changed to accommodate both the new g-plan
structure as well as the firing of e-plans (i.e., plans for reacting
to external events) for all intentions rather than creating a single
new separate intention as before. First, the redefined functions now
receive a set of intentions as an extra parameter; in the new semantic
rules they are called with the agent's plan library and the current
contents of the set of intentions ($\AGPLANS$ and$\CI$ in the
operational semantics, respectively). If $\TE$ is an external event,
the new parameter is used to check for applicable plans for each
individual intention plus the empty intention\footnote{The empty
  intention being included in the set of intentions in that parameter
  is useful for backwards compatibility with traditional AgentSpeak(L)
  but we do not discuss this here as it would hinder the explanation
  of the essential aspects of the proposed extension.}
$\EXT$. Besides having an extra parameter, the functions now return a
set of triples rather than a pair. We use $\TREV(p)$ to refer to the
triggering event of a plan $p$, and $\SCOPE([\PL_1,\ldots,\PL_n])$
returns all plans in the scope of the g-plan $p_n$. By ``in scope'',
we mean the plans that appear (immediately) within the braces
delimiting a g-plan (but not within its subgoals); for example the
plans for \texttt{+e1}, \texttt{+e2}, and \texttt{+!k} (but not for
\texttt{+e3}) are in the scope of the g-plan \texttt{+!g : c <: gc \{
  $\ldots$ \}} shown in the beginning of
Section~\ref{sec:infSS}. Furthermore, note that the $\SCOPE$ function
can determine the exact plan for $\PL_n$ in the forest of g-plan trees
now forming the plan library by using plans
$\PL_1,\ldots,\PL_{n-1}$. We can now formally define $\RELPLANS$:

\begin{definition}[Relevant Plans]\label{def:relplans}
  The auxiliary function to retrieve relevant plans given a plan
  library $PL$, a particular event $\TE$ of type \texttt{+l} or
  \texttt{-l} (i.e., an external event, one reacting to changes in
  beliefs rather than a goal adoption event), and a set of intentions
  $I$ is defined as
  $\RELPLANS(PL,\TE,I) = \{ \langle\PLANS,\theta,i\rangle \mid i \in
  I, i = [\PL_1,\ldots,\PL_n],$
  and
  $ps = \{ rp \mid rp \in \SCOPE([\PL_1,\ldots,\PL_j]) \mbox{ and }
  \{\TE\} \models \TREV(rp)\theta$,
  for some m.g.u.\ $\theta\}$, where $j$ is the greatest number in
  $[1..n]$ s.t.\ $ps \neq \emptyset$, or $ps=\emptyset$ if there is no
  such $j\}$. For internal events, the function returns a set
  containing a single element $\langle\PLANS,\theta,i\rangle$ where
  $i$ is the specific intention that generated $\TE$ and $\PLANS$ is
  the set of relevant plans using the same idea of scope as above for
  external events.
\end{definition}

We do not formally define the $\APPLPLANS$ function here due to space,
and given that it is trivially extended in a very similar way as in
Definition~\ref{def:relplans} for $\RELPLANS$.

Finally, we need to change the semantics so that not just one but
\emph{all} relevant e-plans are triggered, that is, one for
each active intention, and within a single intention starting from the
most specific plan (i.e., the one closest to the top of the intention
if there are relevant plans at other levels too). Most of the work was
already done in the redefinitions of the $\RELPLANS$ and $\APPLPLANS$
functions, but we still need to change the rule for processing
external events (which now change multiple intentions). Note however
that by the time we come to handle an external event with rule
\rn{ExtEv}, a single plan for each intention has already been
selected; that is, $\SO$ (the option selection function) is also
slightly redefined to work with the new structures returned by those
redefined auxiliary functions.

The new \rn{ExtEv} rule is below, and it essentially says that an
external event now potentially interrupts various intentions, provided
there are relevant and applicable plans anywhere in the g-plans
associated with each particular current intention of the agent. The
$\TRHO$ component of the transition system configuration has the
result of the application of the new $\RELPLANS$ and $\APPLPLANS$
functions.
% Recall that external events for the ``main'' goal --- the external
% events as in traditional AgentSpeak --- will now be associated with
% the empty intention \EXT in \TRHO (i.e., the result of the
% $\RELPLANS$ and $\APPLPLANS$ functions).
The chosen plan for each intention, after being applied its respective
variable substitution $\theta$, is pushed on top of that intention
(the notation used for this below is $i_j[\PL_j\theta_j]$).

\infrule[ExtEv] {
        \TEPS = \langle\TE,\EXT\rangle \qquad \TRHO =
        \{\langle\PL_1,\theta_1,i_1\rangle,\ldots,\langle\PL_n,\theta_n,i_n\rangle\}
}
%------------------------------------------------
{   \CFG{\ClrInt}  \trans  \CFGcp{\ClrInt} \\[1.2mm]  
\begin{array}{llcl}
  \mbox{\emph{where:}\quad}
     & \CIli & = & \CI \setminus \bigcup_{j=1}^{n} \{i_j\} \cup
                   \bigcup_{j=1}^{n} \{i_j[\PL_j\theta_j] \}\\
\end{array}
}

To conclude this section, we emphasise that, for the sake of space, we
here formalised only the main changes to the well-known AgentSpeak
semantics. The complete new transition system giving semantics to
AgentSpeak(ER) will appear in a longer paper.

### 2019-06-01 future.tex
first structure

%% RHB: OK to change this title as below?
%\section{Conclusion and Future Work}
\section{Future Work}
\label{sec:future}

\todo{OLD CONCLUSION SECTION}

As with any new programming language, there is much future work, some
in fact ongoing. We are currently refining the ASTRA implementation,
trying to make a few optimisations to improve the evaluation results we
reported in this paper. A Jason-based implementation is also under
way; comparison of the performances of the two implementations might
lead to insight that might improve the implementation of the platforms
themselves.

More generally, full understanding and evaluation of a programming
language takes many years. We expect in the long term to use \aser\ in
the practical development of multi-agent systems, both for real-world
systems and also academic ones (e.g., for the multi-agent programming
contest~\cite{Albrecht18}). However, besides the actual programming
practice, we expect \aser\ to contribute to formal work as
well. Assessing how formal verification of \aser\ systems compares to
the original language is also planned as future work.
%% RHB: OK to mention the multi-agent contest?

% \begin{itemize}
% \item refining current ASTRA implementation, that will allow to refine the evaluation
% \item Jason extension implementing {\aser}, and compare it with the ASTRA one
% \item stress {\aser} in practice 
% \item think about the tools: how existing IDE and tools can be extended to provide functionalities exploiting the new features
% \item investigate how the extension either improve or not the possibility to formally analyse the correctness properties
% \item ...
% \end{itemize}

### 2019-06-01 information-sheet.tex
first structure

\documentclass[11pt]{article}
\usepackage{times}
\title{Information Sheet for `Agent Programming in the Cognitive Era' (Viewpoint Paper)}
\date{}

\begin{document}
\maketitle

\noindent\textbf{What is the novelty and significance of the area being discussed in the context of autonomous agents and multiagent systems?}

The paper explores possible future approaches to the development of autonomous intelligent agents and multiagent systems. Specifically, it examines the claim that in the so-called `Cognitive Era', intelligent systems will be trained using machine learning techniques, rather than developed by human programmers, see, e.g, (Kelly and Hamm, 2013). The means by which agent systems are developed is a central concern of the AAMAS community, and, in particular, of the well-established agent programming subcommunity within AAMAS.

\medskip\noindent\textbf{Why is it timely to provide a viewpoint article in this way now?}

The claim that future intelligent systems will be trained using machine learning techniques rather than developed by human programmers has implications for future research directions in agent programming and the development of methodologies and tools for programming agents. It is particularly topical, given that the AAMAS 2019 conference will, for the first time, include a special track on `Engineering Multiagent Systems'.

\medskip\noindent\textbf{Explain how you have sought to ensure that your viewpoint is not merely an individual perspective with little sound basis, and provides a synthesis of current thinking in addition to your own perspective so that it provides a rigorous analysis of the area.}

We provide a brief survey of the last 20 years of agent oriented programming and agent development platforms. We also survey in some detail previous work on integrating AI components and techniques into the BDI agent architecture and execution cycle. Our discussion of open research problems and possible future research directions for agent programming is grounded in this analysis.

\medskip\noindent\textbf{Have research, survey or viewpoint papers on the same research area been published and, in particular, have any been published recently?}

To the best of our knowledge, no papers have addressed this research area. 
%
The most closely related recent papers we are aware of are:

\emph{K. Kindriks (2014). The Shaping of the Agent-Oriented Mindset - Twenty Years of Engineering {MAS}, Engineering Multi-Agent Systems - Second International Workshop, {EMAS}
               2014, 1--14, Springer}

\emph{B. Logan (2018). An Agent Programming Manifesto. International Journal of Agent-Oriented Software Engineering, 6(2), 187-210.}

\medskip\noindent\textbf{What are the main differences between your viewpoint and these previously published papers?}

The (Hindriks 2014) paper traces the development of the agent oriented approach to programming from the 1990s to the present day. It briefly discusses AI-related topics but does not provide an analysis of the role of machine learning or AI in the future of agent programming research as in the present paper.
%
The (Logan 2018) paper presents an analysis of the limited adoption of agent programming techniques by the mainstream programming community. While it mentions the possible role of AI in the future development of agent oriented programming, it does not survey previous work in this area or offer a detailed discussion of open research problems and future research directions for the integration of AI into the BDI model as the current paper does. It does not consider the question of whether agents can trained rather than programmed.

\medskip\noindent\textbf{Have you published parts of your paper before?}

No.

\end{document}

### 2019-06-01 intro.tex
first structure

\section{Introduction}
\label{sec:intro}

\todo{[OLD VERSION]}

%
% About AgentSpeak(L)
%
{\asl} has been introduced in \cite{Rao96}  with the purpose of defining an expressive, 
abstract language capturing the main aspects of the Belief-Desire-Intention architecture~\cite{Bratman88,Georgeff:1987:RRP:1863766.1863818}, featuring a formally defined semantics and an abstract interpreter.
%
The starting point to define the language were real-world implemented systems, namely the
Procedural Reasoning System (PRS)~\cite{Ingrand:1992:ARR:629535.629890} and the Distributed Multi-Agent Reasoning System (dMARS).
%

%
% Concrete APL
% 
{\asl}  and PRS have become a main reference for implementing concrete Agent Programming Languages 
based on the BDI model: 
%
main examples are Jason~\cite{jason06,bordini:07} and ASTRA~\cite{DBLP:conf/prima/CollierRL15}.
%
Besides Agent Programming Languages, the {\asl} model has been adopted
as the main reference to development several BDI agent-based
frameworks and technologies~\cite{BordiniMAPlpa,BordiniMAPlta} as well
as serving as inspiration for theoretical work aiming to formalise
aspects of BDI agents and agent programming
languages~\cite{DBLP:conf/kr/WinikoffPHT02,DBLP:conf/promas/DennisFBFW07,DBLP:journals/aamas/BordiniFVW06}.

%
% Contribution
%
Existing Agent Programming Languages extended the language with
constructs and mechanisms making it practical from a programming point
of view~\cite{jason06}.
%
Besides, proposals in literature extended the model is order to make
it effective for specific kinds of systems --- e.g., real-time
systems~\cite{Vikhorev:2011:APP:2030470.2030529} -- or to
improve the structure of programs, e.g. in terms of
modularity~\cite{Madden2010,Nunes2014}.
%%RHB: Not sure what should go in XXX above?

%
Along this line, in this paper we describe a novel extension of the
{\asl} model --- called {\aser} --- featuring \emph{plan
  encapsulation}, i.e. the possibility to define plans that fully
encapsulate the strategy to achieve the corresponding goals,
integrating both the pro-active and the reactive behaviour.
%
% Key points
%
This extension turns out to bring a number of important benefits to
agent programming based on the BDI model, namely:
%
\begin{itemize}
\item improving the overall readability of the agent source code,
  reducing fragmentation and increasing modularity;
\item promoting a more goal-oriented programming style, enforcing yet
  preserving the possibility to specify purely reactive behaviour,
  properly encapsulated into plans for goals;
\item improving intention management, enforcing a one-to-one relation
  between intentions and goals --- so every intention is related to a
  (top-level) goal;
\item improving failure handling, in particular making it easier the
  management of failures related to plans for reacting to environment
  events;
\end{itemize}
%
% \noindent Besides the benefits in terms of agent programming, the approach reduces the gap between the design level and the programming level,  ...
% \item facilitate goal-based reasoning -- ...
% \end{itemize}

\noindent The remainder of the paper is organised as follows:
%
first we describe in details the motivations that lead to the proposal
of a new AgentSpeak extension (Section~\ref{sec:motivation});
%
then, we introduce and discuss {\aser}, defining the main concepts,
syntax and semantics --- first informally (Section~\ref{sec:proposal})
and then providing the formalisation of some key aspect
(Section~\ref{sec:formalisation}).
%
We discuss then the results of a first evaluation that has been
carried out, based on a prototype implementation extending the ASTRA
platform (Section~\ref{sec:evaluation}).
%
We conclude the paper discussing related work
(Section~\ref{sec:related}) and sketching future work
(Section~\ref{sec:conclusion}).

### 2019-06-01 motivation.tex
first structure

\section{Motivation}
\label{sec:motivation}

\todo{EMAS VERSION}

The main motivation behind {\aser} comes from the experience using
agent programming languages based on the {\asl} model, {\jason} and
ASTRA in particular.
%%
Yet, these issues are relevant for any language based on the BDI
architecture.
%%
%The first issue is about the plan model.

% \subsection{Plan encapsulation}
%
In the BDI model, plans are meant to specify some means by which an
agent can satisfy an end~\cite{Rao96}.
%
In {\asl}, a plan consists of a rule of the kind \textsf{e : c <- b}.
%
The head of a plan consists of a triggering event \textsf{e} and a
context \textsf{c}.
%
The triggering event specifies why the plan was triggered, i.e., the
addition or deletion of a belief or goal.
%
In the following, we refer to plans triggered by event goals
as \emph{g-plans}, and plans triggered by
belief change (including percepts) as \emph{e-plans}.
%
The context specifies those beliefs that should hold given the agent's
current belief base if the plan is to be triggered.
%
The body of a plan is a sequence of actions or (sub-)goals.
%

%
% THE PROBLEM
%
In this approach --- as well as in planning, in general --- the
\emph{means} to achieve a goal (i.e., the plan body) is meant to be
fully specified in terms of the actions the agent should execute and
the (sub-)goals the agent should achieve or test.
%
In practice, when programming such systems, it is often the case that
the strategy (the means) adopted to achieve some goal (the end)
naturally includes reactions --- i.e., reacting to events asynchronously
perceived from the environment, including changes in the beliefs.
% 
This reflects more than just the ability of an agent to change/adapt
its course of actions; it allows the integration of reactivity as a
core ingredient of the strategy to achieve some goal.
%
This revised notion of a plan is not just a programming feature; it
also occurs naturally in human activity. For example, a fisherman with
the goal of catching fish waits for the event of a tug on their line
indicating a fish is on the hook. Reactivity is a key ingredient of many
activities that we perform to achieve specific goals, not only to
handle events that represents errors or unexpected situations (for the
current courses of actions).
%
It follows naturally that this is also an opportunity to extend the
plan model so as to fully \emph{encapsulate} reactions that are
part of the strategy to achieve the goal, as well as the subgoals that
are specific to that particular goal.

The use of e-plans to achieve goals is actually an important
conceptual brick of the {\asl} model.
%
Let us consider the robot cleaning example used to describe plans
in~\cite{Rao96}. 
%
One of the plans is:

\begin{small}
\begin{verbatim}
+location(waste,X) : location(robot,X) & location(bin,Y)
  <- pick(waste); !location(robot,Y); drop(waste).
\end{verbatim}
\end{small}

\noindent That is, as soon as the robot perceives that there is
waste at its location, then it can pick it up and bring it to the bin.
%
This e-plan is an essential brick of the overall strategy to achieve
the goal of cleaning the environment. The problem here is that it is
an implicit rather than explicit goal of the agent (since it is an e-plan, 
it is executed regardless of the agent currently having the goal of keeping the
environment clean). In practice, we adopt a maintenance
goal~\cite{Duff:2006:PMG:1160633.1160817} to clean the environment,
which includes reacting to cleaning up waste when we see it. In the
above program, this notion cannot be represented and remains in the
mind of the programmer/designer; as there is no g-plan for it, there
is no explicit trace in the agent mental structures about this
goal.

This problem can also be illustrated with the following
scenario. Consider an agent that includes a set of plans (a module
written by a third party) to handle social obligations. The module has
several e-plans for different types of obligations:
\begin{small}
\begin{verbatim}
+obligation(Ag,committed(Goal)) : .my_name(Ag)      <- ...
+obligation(Ag,achieve(Goal))   : .my_name(Ag)      <- ...
-obligation(Ag,Goal) : .my_name(Ag) & .intend(Goal) <- ...
\end{verbatim}
\end{small}
If for some reason during its execution an agent decides not to follow
these plans anymore (e.g.\ it chooses to become disobedient), it is
difficult to ``disable'' the behaviour of the above plans. Either
these plans have to be changed to consider a particular state of the
agent or the agent removes all these plans from its plan
library. Neither option is simple to
program. % suitable: the third party code could be not modified; we do not have references for these plans to remove them.
Although we can solve the problem, the lack of an explicit goal
stating that the agent intends to be obedient is the cause of this
problem.

%
%
%
Besides maintenance goal, also for achievement goals we see benefits in
encapsulating the reactive behaviour in the corresponding plans.
%
Let's consider, as a very simple example, yet capturing the point,  
the goal of printing down all the numbers between N and 1,
stopping if/when a 'stop' percept is perceived.
%
In {\asl} this task can be effectively tackled using only g-plans:
%
\begin{small}
\begin{verbatim}
+!print_nums(0).
+!print_nums(_) : stop.
+!print_nums(N) : not stop <- println(N); !print_nums(N-1).
\end{verbatim}
\end{small}
%
\noindent The action \texttt{println} is meant to print the number
on standard output.
%
Here we exploit the fact that the BDI reasoning cycle automatically 
updates beliefs from percepts, and this allows us to write down 
structured plans with courses of actions that change according 
to the environment, by exploiting goals/subgoals and contexts.
%
%This is a clean solution, exploiting the context in plans
%and the fact that in the reasoning cycle beliefs are automatically
%updated given new percepts.
%
More generally, in {\asl} the suggested approach to realise
structured activities whose flow can be environment dependent is by
means of (sub-)goals and corresponding plans with a proper context.
%
In some cases however this approach is not fully effective, and
\emph{e-plans} are needed,
%
in particular every time we need to react \emph{while executing the
  body of a plan}.
%
A typical case occurs when we have actions (or subgoals) in a plan
that could take a long time to complete.
%
For instance, suppose that instead of simply printing we have a
long-term \texttt{elab} goal (it could be even an action):

\begin{small}
\begin{verbatim}
+!print_nums(0).
+!print_nums(N) : stop. 
+!print_nums(N) : not stop <- !elab(N); !print_nums(N-1).
\end{verbatim}
\end{small}

\noindent Suppose that, realistically, we cannot spread/pollute plans
about the goal \texttt{!elab} with a stop-dependent behaviour.
%
To solve this problem, using {\asl} and in BDI architectures
an e-plan can be introduced, using e.g. internal actions to act on the
current ongoing intention:

\begin{small}
\begin{verbatim}
+!print_nums(0).
+!print_nums(N) <- !elab(N); !print_nums(N-1).		
+stop <- .drop_intention(print_nums(_)).
\end{verbatim}
\end{small}
%% RHB: should we change .abort to .drop_intention ?

\noindent The problem here is that the e-plan \texttt{+stop <- ... } is
part of the strategy to achieve the goal \texttt{!print\_nums}, however
it is encoded as a separate unrelated plan.

Finally, the encapsulation of  also impacts  plan failures handling, which is a very
important aspect of agent programming.
%
In the Jason dialect of {\asl}, we can define plans that handle the
failure of the execution of g-plans (generating the event
\texttt{-!g}), but not of e-plans.
%
For example, if there is a problem in the println action in:

\begin{small}
\begin{verbatim}
+stop <- println("stopped").
\end{verbatim}
\end{small}

\noindent the plan execution fails, without any possibility to react
and handle the failure.
%
In order to handle this, a programmer is forced to structure every
e-plan with failure handling using a subgoal:

\begin{small}
\begin{verbatim}
+stop <- !manage_stop.
+!manage_stop <- println("stopped").
-!manage_stop <- println("failure").
\end{verbatim}
\end{small}

\noindent This contributes to making the program longer and verbose,
besides increasing the number of plans to be managed by the
interpreter.

%Besides failure, meta-level actions to manage intentions (plans in execution) at runtime
%works only with g-plans, since they need to specify as argument the specific goal which generated the intention.

\bigskip

To summarise, this is the set of key issues identified for the basic
plan model in the practice of agent programming:
%
\begin{description}
%
\item[Lack of encapsulation:] The strategy to achieve a goal is
  fragmented among multiple plans, not explicitly related to each
  other.
%
\item[Implicit vs. Explicit goals:] Some parts of an agent program may
  have plans for which the goal is implicit.
%
\item[Difficult failure handling:] Failures/errors generated in the
  body of e-plan cannot be directly captured.
%
\end{description}
%%RHB: Are we not talking about the goal-condition and the fact that
%%external events can trigger various plan, one for each intention??

%\noindent This has an impact also on the performance of the agent
%interpreter at runtime. In the traditional model, each reaction
%represented by an e-plan necessarily generates a new intention to
%manage it, increasing the number of plans in execution to be managed
%by the agent reasoning cycle.
%
%This could have also a negative impact on intention selection at
% each reasoning cycle~\cite{DBLP:conf/atal/LoganTY17}.

%\item[Difficult runtime management] -- Intentions generated by e-plans cannot be managed. Besides, as a consequence of point (1), it is not possible to manage (suspend, abort) the intentions related to some goals as a whole.
%
%\item[Runtime overhead] Each reaction (e-plans) is   new intention => new stack.

\noindent Besides the pure programming perspective, it is worth noting that
these issues can affect the %whose
software engineering process for agent development.
%%
In particular, at the design perspective, it is natural to specify
coarse-grained plans fully encapsulating the strategy to achieve or
maintain goals.
%
It would be important then to keep as much as possible the same level
of abstraction when going from the design to programming, and at
runtime too, to support agent reasoning.

%
%\begin{itemize}
%%
%\item Design / Engineering Drawbacks -- weak encapsulation => impact on modularity => impact on many aspects
%
%%
%\item Reasoning drawbacks -- proliferation of intentions => impact on the reasoning cycle, on intention selection
%%
%\end{itemize}

### 2019-06-01 proposal.tex
first structure

\section{The {\aser} Proposal}
\label{sec:proposal}

\todo{OLD VERSION}

To overcome the problems discussed above, we consider two key changes
in the plan model.
%
The first one is to extend the plans beyond the simple sequence of
actions and goals, so as to include also the possibility to specify
reactive behaviour encapsulated within the plan.
%
Coherently, with the {\asl} model, such a behaviour can be expressed
in terms of e-plans.
%
% JH: I changed a bit the terminology
Accordingly, a plan becomes the \emph{scope} of ($i$) a sequence of
actions (referred as \emph{body actions}), ($ii$) a set of
\emph{e-plans}, specifying a reactive behaviour which is active at
runtime only when the plan is in execution, and ($iii$) a set of
\emph{g-plans}, specifying plans to achieve subgoals that are relevant
only in the scope of this plan. The e-plans and g-plans are referred
to as \emph{sub-plans}.
%
The sub-plans may include also reactions to failures occurring when
the plan is executed.
%% RHB: So the failure handling plans for the outer goal were moved
%% inside (as well as out)? I'm against this, but I guess this is a
%% democracy :D

%
%
%
The second one is enforcing that e-plans --- as pieces of reactive
behaviours --- must always be defined in the context of a g-plan. We
are thus enforcing the principle that an agent does (and reacts)
always because of a goal to achieve or maintain.
%
This ensures that programmers explicitly specify what is the goal to
achieve even when defining a purely reactive behaviour.
%
In so doing, at runtime every intention\footnote{As in \asl, an
  intention is the result of the deliberation to commit to some
  desire. Briefly, if the agent has an applicable plan for a goal
  event (i.e.\ a desire), it commits to it by creating an intention
  based on that plan and starts executing it.} has an
associated goal being pursued.
%

In the remainder of the section, we first describe in detail the
syntax and informal semantics of the new plan model, including simple
examples, and then we discuss the key benefits.

\subsection{Informal Syntax and Semantics}
\label{sec:infSS}

At the top level, an {\aser} program is a collection of g-plans, whose
syntax is exemplified below:
%
% CURRENT VERSION 0.2 (in the doc)
%
{\small
\begin{verbatim}
/*  g-plans to achieve goal g in context c */
+!g : c <: gc {

  <- a; b; ?g1; !g1; !!g2. // plan body (optional).

  /* e-plans */
  +e1 : c1 <- b1.	
  +e2 : c2 <- b2.
  +!k : true <: b(10) {
    <- a,b,c.
    +e3 : c3 <- b3. // possible old-style plans
  }

  /* e-plans catching failures */
  -!g[error(ia_failed)] : ...
    <- ... catches from failures

  /* further g-plans */
  +!g1 : c1 <: gc1 { ... }
}
\end{verbatim}}

Like in {\asl}, a g-plan is defined with a head and a body.
%
% JH: avoid to use the term intention here... goal oriented
% programming. explain intention in the semantics 
Besides the triggering event and the context, in \aser\ the head has a
third new element: a \emph{goal condition}, optionally written after
\texttt{<:}, with the same syntax as the context. While the context is
a pre-condition to select a plan as applicable for an event, the goal
condition is a post-condition that defines when the goal can be
considered as achieved\footnote{Or considered impossible to achieve,
  or the motivation for the goal no longer holds, etc. Programmers can
  use this for any condition that implies the goal should no longer be
  pursued. Note that, when programming declaratively, this condition
  is likely to include the goal in the triggering event itself (as
  believed to be true).}. Any goal created based on this g-plan is
considered achieved if and only if
%% RHB: isn't if and only if too strong?
%% Jomi is probably right, keeping if and only if
this condition holds. If no goal condition is specified, the goal is
considered achieved as soon as the body execution completes, as usual
in \asl. However, if a goal condition is defined, having finished the
body execution is not sufficient to deactivate the goal. Notice that
if the goal condition becomes true while the body actions are being
executed, the execution ceases immediately, since the goal being
achieved means no further action would be necessary.
% aborted. %%AR: isn't 'aborted' suggesting a kind of failure? 
% JH: interrupted is also not but, since it will not be resumed, I changed to stop

In {\aser} the body defines the g-plan scope enclosed by `\texttt{\{}'
and `\texttt{\}}' and is composed of the body actions (after `\texttt{<-}'
and before `\texttt{.}')  as well as sub-plans.
%
Like in {\asl}, as soon as a g-plan is instantiated, the body actions
start to execute. The body actions can be empty --- this is the case
of g-plans expressing a purely reactive behaviour.
%
The sub-plans of the g-plan are considered as relevant only for events
produced by the g-plan execution, since they are in the scope of the g-plan.

% GoalCondition is analogous to the context: however, if context can be considered to be as the pre-condition to execute the plan, the GoalCondition are like a post-condition, which corresponds to a declarative description of the goal to be achieved. 
%
%Alternatively, the execution of a plan can be forced to complete by using an internal action .done - this is useful everytime the goal condition cannot be effectively or naturally described as a boolean expression.
%
%If [GoalCondition] is not specified, by default the condition is that all actions of the main sequence have been executed and completed, in continuity with the AS model. This condition is explicitly represented by the new predefined predicate \texttt{.is\_done}.
%
% The body of the plan provides a lexical and runtime scope of the sub-plans, that is:  variables used in the g/c expression are visible also to subplans the lifetime/availability of the sub-plans is limited to the time in which the g-plan is in execution

While a g-plan is executing, it can be \emph{interrupted} by events
relevant for its e-plans. When the agent perceives an event and an
e-plan from g-plan is applicable according to its specified context,
then the execution of the g-plan body is suspended until the body of
the e-plan finishes its execution.
%
In {\asl}, other plans do also interrupt the execution of a plan in
the case of subgoals. For example, in the body \textsf{a1; a2; !g1;
  a3}, \textsf{g1} is a sub-goal and thus the action \textsf{a3} is
executed after the body of \textsf{g1} has finished. The plan body
execution in interrupted synchronously.
%
In {\aser} this uniformly occurs also with any kind of events, not
only sub-goal events; that is, the body actions can be interrupted to
react to events coming from the environment. However, in this case the
interruption is \emph{asynchronous} -- the point where the body
actions is interrupted is unknown, and depends on the environment, at
runtime.
%
This behaviour is ruled by the (extended) reasoning cycle, which will
be described in next section.

% Failures generated by either the main sequence or by sub-plans generate a -!g that could handled.
%[TO BE COMPLETED?]
%
\subsection{Examples}
To give a more concrete taste of the language, we consider again the
examples seen in Section~\ref{sec:motivation}, now rewritten in
{\aser}.
%
The robot cleaning example becomes:

\begin{small}
\begin{verbatim}
+!clean_env <: false {
   +location(waste,X) : location(robot,X) & location(bin,Y)
      <- pick(waste); !location(robot,Y); drop(waste).
}
\end{verbatim}
\end{small}

\noindent We can give an explicit reason for the reactive behaviour by
encapsulating the e-plan inside a g-plan, with an explicit goal
\texttt{clean\_env}.
%
Setting the goal condition to \texttt{false} means that the plan
execution is going to last forever. It is a way of implementing some
form of maintenance goal.
%
This is also a particular case where the body of the g-plan does not
have any actions.
%

The \texttt{elab\_nums} example seen before can be rewritten to fully
encapsulate the strategy in the same g-plan:

{\small
\begin{verbatim}
+!elab_nums(0).
+!elab_nums(N) <: stop | .body_done {
   <- !elab(N); !elab_nums(N-1).	
   +!elab(M) <- ...
}
\end{verbatim}}

  \noindent Here \texttt{.body\_done} is a predefined internal action
  that succeeds when the body actions have been completed. The goal
  \texttt{elab\_nums} is achieved either by the perception of
  \texttt{stop} or by the execution of its body actions.
%
  In case we want to take some action to react to \texttt{stop}, we
  can introduce an e-plan as follows:

{\small
\begin{verbatim}
+!elab_nums(0).
+!elab_nums(N) {
   <- !elab(N); !elab_nums(N-1).
   +!elab(N) <- ...
   +stop <- println("stopped"); .done.
}
\end{verbatim}}

  For the example of plans to handle social obligations, we can define
  an explicit goal by encapsulating the e-plans in a g-plan as
  follows:
\begin{small}
\begin{verbatim}
+!be_obedient <: false {
   +obligation(Ag,committed(Goal)) : .my_name(Ag)      <- ...
   +obligation(Ag,achieve(Goal))   : .my_name(Ag)      <- ...
   -obligation(Ag,Goal) : .my_name(Ag) & .intend(Goal) <- ...
   ...
}
\end{verbatim}
\end{small}
\noindent Now the agent can disable these plans by simply performing
\texttt{. drop\_intention(be\_obedient)}, and resuming its obedient
behaviour by adopting the goal \texttt{!be\_obedient}, and it can also
check whether it is being obedient by testing
\texttt{.intend(be\_obedient)}.

With the new language, we can program different kinds of commitments
to goals~\cite{cohen:90,winikoff:02} by exploiting the goal
condition. For instance, the Single Minded Commitment for goal
\texttt{g} can be programmed as:
\begin{small}
\begin{verbatim}
+!g <: g | f {
   ...
}
\end{verbatim}
\end{small}
where \texttt{f} states when the goal becomes impossible. The agent
commits to achieve \texttt{g} until \texttt{g} is believed either true
or impossible to achieve. To program this kind of commitment in \asl,
we have to follow some programming patterns requiring extra plans
(three extra plans as shown in~\cite{hubner:06b}).

%%%RHB: The paragraph below isn't clear to me
%  \bigskip The proposed approach allows us to avoid towers of subgoal
%  calls, that are used when a plan does some action and then should
%  wait that some condition is achieved, in order to complete,
%  eventually reacting to some events to adapt the course of actions.
%%
%An example is a simple thermostat. 
%%
%Let's consider the goal to achieve some temperature +!temp(T):
%
%{\small
%\begin{verbatim}
%+!temp(T) : temp(CT) & CT < T & not warming
%  <-  switchOnWarming; !temp(T).
%+!temp(T) : temp(CT) & CT > T & not cooling
%  <-  switchOnCooling; !temp(T).
%+!temp(T) : temp(CT) & CT != T 
%  <-  !temp(T).
%+temp(T) : temp(T) <-  stop.
%\end{verbatim}}
%
%  \noindent There is here a kind of \emph{cognitive busy waiting}
%  situation: after switching on the HVAC system (to cool down or warm
%  up), we need to check continuously the condition to
%  know when/if the goal has been achieved, before ending the plan.
%%
%  This is problematic for 2 reasons: the size of the intention stack
%  is growing indefinitely and performance overhead.
%% JH: not sure we should go for performance arguments as above. the
%% need to check all conditions every cycle is equivalent to the cost
%% of the above cognitive busy waiting
%% RHB: Furthermore, is the intention stack really growing indefinitely
%% in this case??? Probably not as it's just tail recursion and Jomi
%% optimised that in Jason, no need to use !! anymore. 
%%
%The \aser version avoids that problem:
%
%{\small
%\begin{verbatim}
%+!temp(T) : temp(CT) & CT < T 
%  <: temp(T) & not warming & not cooling {  
%  <- switchOnWarming.
%  +temp(T) <- stop. // There is bound
%}
%
%+!temp(T) : temp(CT) & CT > T 
%  <: temp(T) & not warming & not cooling {  
%  <- switchOnCooling.	
%  +temp(T) <- stop.		
%}
%+!temp(T) : temp(T).
%\end{verbatim}}
%
%% JH: I am not sure to present this example, seems more complex than original code
%% JH: it is not equivalent to ASL version: if the first plan is
%% triggered, how could the second be triggered latter?
%% RHB: Personally, I don't like examples about thermostats very much...
%
%Remarks:
%\begin{itemize}
%\item the T variable used inside the plan body (+temp(T) <-..) has been bound in +!temp(T)
%\end{itemize}
%
%\noindent  In Jason, the original AgentSpeak(L) solution could be avoided  by exploiting a .wait internal action that suspends a plan until some condition is achieved:
%
%{\small
%\begin{verbatim}
%+!temp(T) : temp(CT) & CT < T 
%  <-  switchOnWarming;
%      .wait(temp(T));
%      stop.
%
%+!temp(T) : temp(CT) & CT > T 
%  <-  switchOnCooling;
%      .wait(temp(T));
%      stop.
%\end{verbatim}}
%
%  \noindent However, this solution is no longer effective (by itself)
%  if something could happen while waiting which would require changing
%  the course of action.
%%
%  For instance, suppose that while warming up, the temperature value
%  surpasses the target value T, such that we would need to start
%  cooling. In this case, the .wait solution must be strongly reworked
%  and becomes complicated.
%%
%The complete solution using the extended plan model is as follows:
%
%{\small
%\begin{verbatim}
%+!temp(T) <: temp(T) & not warming & not cooling {
%  +temp(T) <- stop.		
%  +temp(T1) : T1 > T <- switchOnCooling.
%  +temp(T1) : T1 < T <- switchOnWarming.
%}
%\end{verbatim}
%}
%
%% JH: shouldn't we present only the above solution? clear and simpler 
%% is the test not warming/ not cooling necessary on the GC
%
%% RHB: The goal condition of the plan above is strange. What is if
%% for? Isn't it the same as .done in the first +temp plan?
%
%\noindent Maintenance goal: to achieve and keep the temperature at the target temperature T, with the possibility of dynamically updating/changing T.
%% RHB: The phrase above needs improvement
%
%{\small
%\begin{verbatim}
%+!target_temp(T) <: false {
%  <- +target_temp(T).	
% 
%  +temp(T) : target_temp (T) <- stop.		
%  +temp(T1) : target_temp(TT) & T1 > TT <- switchOnCooling.	  
%  +temp(T1) : target_temp(TT) & T1 < TT <- switchOnWarming.
% 
%  +target_temp(TT) : temp(TT)  <- stop.	
%  +target_temp(TT) : 
%    temp(T) & T < TT & not warming <- switchOnWarming.
%  +target_temp(TT) : 
%    temp(T) & T > TT & not cooling <- switchOnCooling.
%}
%\end{verbatim}}

 \subsection{Failure Management}

 Failures generated by the execution of actions of sub-plans for a
 goal \texttt{g} can be handled and managed by:

%%RHB: Carefull, changed but feel free to 
\begin{itemize}
\item subplans of type \texttt{-!g} listed in the body of the g-plan;
\item plans at the same level of \texttt{+!g}, if the event is not
  managed within the body.
\end{itemize}

\noindent In terms of style, it is more in keeping with the new
language to include the failure plans within the scope of the
g-plan. However, it is also true that if we define a plan for
\texttt{+!g} at a certain level of a g-plan tree\footnote{Note that
  g-plans within g-plans now implicitly form a plan tree for a
  top-level goal, and the plan library is thus a forest of such
  trees.}, it might make more sense to the programmer if the goals for
\texttt{-!g} are all placed at the same level (specially if the
programmer is influenced by the style of the Jason variant of
AgentSpeak).

% \noindent In the latter case, the event representing the failure could
% be exactly \texttt{-!g}, as in Jason, i.e. the event which is
% generated when the goal is removed.
%
% This strategy cannot be adopted in the former case, since we want to
% react to a failure but the corresponding goal and intention has not
% been removed yet, so no -!g event is generated. For this reason, we
% need to introduce a new event: \texttt{+/err\_term} [TEMPORARY SYNTAX]
% representing the failure of an action belonging to the main sequence
% or to the body of subplans.
%%
%% RHB: I find this rather clumsy, besides not being formalised. Is it
%% worth going into details of a new mechanism for plan failure? We
%% could just claim it's as before plus now they can be inside the
%% g-plan IF YOU MUST :D
%% Besides, I can't see why we can't generate a -!g event even without
%% removing the rest of the intention (that's what Jason always did
%% anyway).

%% RHB: I'm changing this as discussed with Jomi but feel free to
%% bring it back
% {\small
% \begin{verbatim}
% +!g : c <: GoalCondition {
%   <- a; b; ? ; c. 
%   ...
%   +/err_term : cn <- bn
% }
% \end{verbatim}}

%   \noindent In the reasoning cycle, the \texttt{+/err\_term} is
%   treated as the other events: if an applicable plan is found, it is
%   selected and executed (in the same intention).
% %
%   Otherwise, the behaviour is like it is in pure AgentSpeak(L) --- the
%   whole intention is removed and a \texttt{-!g} event is generated.

%\subsection{Discussion}
% [TO BE COMPLETED]
%
% In this section we go back on the key issues, showing how the approach is effective in overcoming them.
%
\subsection{Key points}

Before giving a more formal description of the semantics, we conclude
this section by summarising the key points brought out by {\aser}:
%
\begin{description}
%
\item[Encapsulation] -- The strategy to achieve a goal is encapsulated
  in one or multiple g-plans, each embedding also the reactive
  behaviour which is part of the strategy.
%
  The effect is to reduce code fragmentation, improving its
  understanding.
%
\item[Explicit goals] -- Every behaviour of the agent is now
  explicitly related to some goal to achieve.
%
  This promotes a more goal-oriented programming style, yet preserving
  the possibility to easily define g-plans based on purely reactive
  strategies;
%
  this allows the agent to better \emph{manage} its intentions. For
  instance, a programmer can now do a simple
  \texttt{.drop\_intention(g)} to disable the behavior of e-plans
  embedded in the corresponding g-plans.
%
% In {\asl} this is quite tricky to do.
In {\asl}, the relation goal-intention is not one-to-one. {\asl} can have intentions from e-plans and so intentions without a explicitly represented goal. 
In {\aser} there is a one-to-one relation between goals and intentions. 
%
All intentions come from goals (only explicit goals are allowed). 
%
Primitives to handle goals can thus handle \emph{all} intentions. 
%
In {\asl}, primitives to handle goals can manage a limited set of the
intentions (those created from goal addition plans).

\item[Failure handling] -- Thanks to explicit goals, failures
  generated in the body of an e-plan can now be directly captured and
  managed by failure recovery plans defined for g-plan, without the
  need for auxiliary goals and plans to be introduced.

\item[Coarse-grained intentions] -- In {\asl}, each e-plan in
  execution has its own intention/stack, which runs concurrently to
  the other intentions. Conceptually, this follows the idea that
  the management of environment events are not part of an existing
  overall plan to achieve some goal.
%
  In the new model instead, an e-plan inside a g-plan is meant to
  specify a behaviour that is useful for achieving the goal of the
  g-plan, so part of the same intention.
%
  For this reason, if an e-plan (sub-plan) inside a g-plan in
  execution is triggered, no new intentions are generated and the body
  of the sub-plan is placed on the top of the stack of the same
  intention.
%
  The new model leads then to more coarse-grained intentions.
%
%%% JH: commented for now, while we are discussing it
%The general effect is to reduce the number of intentions. 
%
%This has an impact on the reasoning cycle, for instance on the step selecting the intention to carry on.
\end{description}
%\subsubsection{Intention implicit Interruption}
%
%The effect is to implicitly interrupt and suspend the main sequence (body) of the g-plan - if available and if not already completed.
%%
%Conceptually, this is the asynchronous version of the suspension that occurs when a subgoal is specified (!g) in the body of a plan. This causes exactly the suspension of the body and the pushing of the body of the new plan on the top of the stack of the intention.

% \subsubsection{Compatibility with {\asl}}

### 2019-06-01 related.tex
first structure

\section{Related Work}
\label{sec:related}

\todo{OLD}

{\aser} is primarily related to work in literature focusing on improving cognitive BDI agent programming.
%
% MODULARITY
%
A main aspect widely discussed and developed in the literature is  \emph{modularity}~\cite{Madden2010, Busetta2000,5285116,Novak:2006:MBA:1160633.1160814,Ortiz-Hernandez2016,vanRiemsdijk:2006:GMA:1160633.1160864,Hindriks2008,Nunes2014}.
%
In programming languages, modularity is strongly related to encapsulation, in fact  strengthening encapsulation typically leads to refining modularity, in particular devising more coarse-grained modules.
%
{\aser} enriches the spectrum of approaches elaborated in literature for improving modularity in BDI-based agent programming languages by devising coarse-grained plans as modules encapsulating goal-oriented \emph{and} reactive behaviour.

Besides, {\aser} is related to existing BDI agent programming languages extending the basic plan model as found in the original proposal of {\asl}.
%
In this context, a main reference is \textsf{CANPlan}~\cite{Sardina2011}, a  BDI-style agent-oriented programming language enhancing usual BDI programming style with declarative goals, look-ahead planning, and failure handling. 
%
It allows programmers to mix both procedural and declarative aspects of goals, enabling reasoning about properties of goals and decoupling plans from what these plans are meant to achieve. 
%
The lookahead planning makes it possible to guarantee goal achievability and avoid undesirable situations. 
%
The plan model adopted in {\textsf{CANPlan}} is analogous to the {\asl} one. Each plan is characterised by a plan rule \textsf{e(t) : $\psi$(xt, y) $\leftarrow$ P(xt, y, z).}, where \textsf{P} is a ``reasonable strategy'' to follow when \textsf{$\psi$} is believed true in order to resolve/achieve the event.
%
\textsf{P} can be a rich composition of actions but not reactions.  
%
Reactive behaviours can be expressed instead --- like in {\asl} and in the basic BDI --- as separate plans handling belief updates corresponding  to environment events.

### 2019-10-01 formalisation.tex
added ASER grammar

@@
-\section{Formal Semantics}
+\section{Formal Syntax and Semantics}
 \label{sec:formalisation}
+
+We first give the formal syntax of \aser, as per the following
+grammar:
+
+\begin{figure}[htbp]
+  \footnotesize{\ttfamily
+    \begin{center}
+      \begin{tabular}{l c l l} 
+   ag & ::= & bs gs gps \\
+   bs & ::= & af$_1$ ... af$_{\textrm{n}}$   & (n $\geq$ 0)\\
+   gs & ::= & \textbf{!}af$_1$ ... \textbf{!}af$_{\textrm{n}}$ & (n $\geq$ 0)\\
+   gps & ::= & gp$_1$ ... gp$_{\textrm{n}}$  & (n $\geq$ 1)\\
+   gp & ::= & \textbf{+!}af \textbf{:} lf \textbf{<:} lf \textbf{\{} [
+              \textbf{<-} h\textbf{.} ] ( gp \galt op )* \textbf{\}}
+              \\ % g-plans 
+   op & ::= & te \textbf{:} lf \textbf{<-} h\textbf{.} \\ % old style plans
+   te & ::= & \textbf{+}af \galt \textbf{-}af \galt \textbf{+}g  \galt
+              \textbf{-}g \\ 
+   g  & ::= & \textbf{!}af \galt \textbf{?}af \\
+   af & ::= & P\textbf{(}t$_1$\textbf{,} \ldots \textbf{,}
+              t$_n$\textbf{)} & (n $\geq$ 0)\\   
+   lf & ::= & af \galt $\neg$ lf \galt (lf $\wedge$ lf) \galt (lf $\vee$
+              lf) \galt \texttt{true} \\
+   h  & ::= & h$_1$\textbf{;} \texttt{true} \galt \texttt{true} \\
+   h$_1$ & ::= & a \galt g \galt h$_1$\textbf{;} h$_1$ \\
+   a  & ::= & A\textbf{(}t$_1$\textbf{,} \ldots \textbf{,} t$_n$\textbf{)} &  (n $\geq$ 0) \\
+\end{tabular}
+    \end{center}
+}
+\caption{Syntax of \aser}
+\label{fig:syntax}
+\end{figure}
+
+\noindent
+where metavariable $P$ stands for a predicate symbol, $t_i$ stand for
+first order terms (as in Prolog for example), and metavariable $A$
+stands for an action symbol. Note also that this is an abstract
+syntax, it contains only the fundamental aspects of the language for
+the purposes of formalisation. Those familiar with the actual systems
+where \aser has been implemented should be aware that all the
+syntactic features of the original systems are of course kept. For
+example, the syntax here does not include the \verb|!!| operator,
+annotations, and many other \jason features.
 
@@
 
-Briefly, these are the changes in the semantics of AgentSpeak(L)
-that are needed to accomodate the new features of AgentSpeaker(ER):
+Briefly, these are the changes in the semantics of \asl
+that are needed to accomodate the new features of \aser:
 

### 2019-10-02 formalisation.tex
cosmetic improvements on ASER grammar

@@
    ag & ::= & bs gs gps \\
-   bs & ::= & af$_1$ ... af$_{\textrm{n}}$   & (n $\geq$ 0)\\
-   gs & ::= & \textbf{!}af$_1$ ... \textbf{!}af$_{\textrm{n}}$ & (n $\geq$ 0)\\
-   gps & ::= & gp$_1$ ... gp$_{\textrm{n}}$  & (n $\geq$ 1)\\
+   bs & ::= & af$_{\texttt{1}}$ ... af$_{\texttt{n}}$   & (n $\geq$ 0)\\
+   gs & ::= & \textbf{!}af$_{\texttt{1}}$ ... \textbf{!}af$_{\texttt{n}}$ & (n $\geq$ 0)\\
+   gps & ::= & gp$_{\texttt{1}}$ ... gp$_{\texttt{n}}$  & (n $\geq$ 1)\\
    gp & ::= & \textbf{+!}af \textbf{:} lf \textbf{<:} lf \textbf{\{} [
@@
    g  & ::= & \textbf{!}af \galt \textbf{?}af \\
-   af & ::= & P\textbf{(}t$_1$\textbf{,} \ldots \textbf{,}
-              t$_n$\textbf{)} & (n $\geq$ 0)\\   
+   af & ::= & P\textbf{(}t$_{\texttt{1}}$\textbf{,} \ldots \textbf{,}
+              t$_{\texttt{n}}$\textbf{)} & (n $\geq$ 0)\\   
    lf & ::= & af \galt $\neg$ lf \galt (lf $\wedge$ lf) \galt (lf $\vee$
               lf) \galt \texttt{true} \\
-   h  & ::= & h$_1$\textbf{;} \texttt{true} \galt \texttt{true} \\
-   h$_1$ & ::= & a \galt g \galt h$_1$\textbf{;} h$_1$ \\
-   a  & ::= & A\textbf{(}t$_1$\textbf{,} \ldots \textbf{,} t$_n$\textbf{)} &  (n $\geq$ 0) \\
+   h  & ::= & h$_{\texttt{1}}$\textbf{;} \texttt{true} \galt \texttt{true} \\
+   h$_{\texttt{1}}$ & ::= & a \galt g \galt h$_{\texttt{1}}$\textbf{;} h$_{\texttt{1}}$ \\
+   a  & ::= & A\textbf{(}t$_{\texttt{1}}$\textbf{,} \ldots \textbf{,} t$_{\texttt{n}}$\textbf{)} &  (n $\geq$ 0) \\
 \end{tabular}
